My title

### 2024-12-02 paper/paper.qmd
finished model and references.bib

library(tidyverse)
library(arrow)
library(here)
library(tibble)
library(knitr)
library(modelsummary)
library(marginaleffects)
library(kableExtra)

@@

All cricket data is from the [@citeR] package [@cricketdata]. The data that was used is from the CRAN version of the package and not the [@github] developer tools version because the CRAN is more stable. [@cricketdata] takes data from [@cricinfo] and [@cricsheet]. While not stated by [@cricketdata], based on the code the team behind the package are scrapping the HTML of ESPN Cricinfo and Cricsheet, As a result, anyone using the package must respect the rate limits set by these sites. Therefore, the data was first saved as CSV before performing any other cleaning and analysis.

More information about the raw data sets, measurements, [@cricsheet], [@cricinfo], and [@cricketdata] is in @sec-data-sheet.

@@

```{r}
#| label: fig-prevoverwickets-wickets
#| fig-cap: Correlation Between Wickets in the Last and Next Over
#| echo: false
#| warning: false
wickets_prev_over_wickets <- cleaned_data %>%
  group_by(over) %>%
  summarise(
    num_wickets = sum(wicket == TRUE),
    prev_over_wickets = mean(prev_over_wickets),
    num_balls = n(),                             
  ) %>% arrange(desc(num_wickets), desc(num_balls))

@@

## Model Set-Up for Single Variable Generalized Linear Model

@@

Beta Explaiation:

$\beta_0$: Represents the general intercept, which represents the log-odds of a wicket occuring when the over variables does not matter

$\beta_1$: Represents the coefficent for the current over number

@@

## Model Set-Up for Two Variable Variable Generalized Linear Model

@@

$\beta_1$: Represents the coefficent for the current over number

$\beta_2$: Represents the coefficent for the number of wickets in the previous over

@@

complex_glm_wicket_model <- readRDS(here("models/complex_glm_wicket_model.rds"))
#summary(complex_glm_wicket_model)
modelsummary(list("Wicket Prediction (Two Variable Input)" = complex_glm_wicket_model))
```
Based on @tbl-model-summary-two-var, modeling with only the over variable accounts 0.22 RSME variability. Also for an increase in number previous wickets in the previous over, the log odds of a wicket occuring increase by 1.852 with a p value that is almost zero (less than $2*10^-16$). The null deviance is 20849 on 51586 degrees of freedom while the residual deviance is 16584 on 51584 degrees of freedom. The logs odds increase for over is 0.005 with a p value of 0.168.

## Model Set-Up for Four Variable Generalized Linear Model

Adding the two extra variables, batting style and bowling style, to the model did not change the effecitveness in predicability at all. More rigouris analysis avaliable in @sec-fourvar-model.

@@

Based on @fig-over-wickets, there was a strong positive linear relationship between the over and the number of wickets taken, suggestign that the over can be a predictor of a wicket occuring. From @fig-bowlingstyle-wickets certain matchup produced a higher chance of a wicket occuring, such as left hand fast on a left hand batter, which is why both variables were added to the four variable model in @sec-fourvar-model. Finally from @fig-prevoverwickets-wickets, the average number of wickets that occured in the previous over had a strong positive linear relationship with the number of wickets that occured in balls in the next over, implying that previous over wickets impact the wicket probability of a ball in the next over.

@@

## Training and Test Split

80% of the data was used for training and 20% of the data will be used for testing.

@@

simple_glm_wicket_model_predictions |>
  mutate(wicket = factor(wicket)) |>
  ggplot(aes(x = over, y = estimate, color = wicket)) +
  stat_ecdf(geom = "point", alpha = 0.75) +
  labs(
    x = "The Over",
    y = "Estimated Probability that a wicket will occur",
    color = "Was actually a wicket"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-model-one-var-results
#| tbl-cap: "One Variable Model Underfits the Data"
#| echo: false
#| warning: false
test_data_simple <- test_data
predictions <- predict(simple_glm_wicket_model, newdata = test_data_simple, type = "response")

@@

summary_results <- test_data_simple %>% group_by(wicket) %>% 
  summarise(
  correct = sum(correct_prediction),
  incorrect = sum(!correct_prediction)
)

summary_results

```

@@

```{r}
#| label: fig-twovarmodel-results
#| fig-cap: Logarithmic Relationship Between Wicket Probability and the previous overs wickets
# (after adjusting for the over)
#| echo: false
#| warning: false
complex_glm_wicket_model_predictions <-
  predictions(complex_glm_wicket_model) |>
  as_tibble()

complex_glm_wicket_model_predictions |>
  mutate(wicket = factor(wicket)) |>
  ggplot(aes(x = prev_over_wickets, y = estimate, color = wicket)) +
  stat_ecdf(geom = "point", alpha = 0.75) +
  labs(
    x = "The Number of Wickets Taken in the Previous Over",
    y = "Estimated Probability that a wicket will occur",
    color = "Was actually a wicket"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-model-one-var-results
#| tbl-cap: "Two Variable Model Struggles to Predict Wickets"
test_data_complex <- test_data
predictions <- predict(complex_glm_wicket_model, newdata = test_data_complex, type = "response")

summary_results <- test_data_simple %>% group_by(wicket) %>% 
  summarise(
  correct = sum(correct_prediction),
  incorrect = sum(!correct_prediction)
)

summary_results
```

@@

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

# Four Variable Model {#sec-fourvar-model}

@@

$\beta_1$: Represents the coefficent for the current over number

$\beta_2$: Represents the coefficent for the number of wickets in the previous over

$\beta_3$: Represents the coefficent for the bowling style. Each bowling style has its own coefficent.

$\beta_4$: Represents the coefficent for the batting style (left handed or right handed). Each batting style has its own coefficent.

@@

```{r}
#| label: tbl-model-summary-four-var
#| tbl-cap: "Batting Style vs Bowling Style Do Not Heavily Impact Wicket Potential"
#| echo: false
#| warning: false
overly_complex_glm_wicket_model <- readRDS(here("models/overly_complex_glm_wicket_model.rds"))
#summary(overly_complex_glm_wicket_model)
modelsummary(overly_complex_glm_wicket_model)
```
Based on @tbl-model-summary-four-var, none of the batting styles or bowling style coefficent are statistically significant all with p values greater than 10%. This implies that most likely these variables not great predictors of wicket creation. The number of wickets in the previous over still has a statistically signifcant impact and has an increase of 0.031446 log odds per an increase of one wicket in the previous over. The residual deviance is 16574 on 51571 degrees of freedom compared to the null of 20849 on 51586 degrees of freedom.

# Variables Not Part of Model

@@

cleaned_data %>% head()

@@

# Extra Tables {#sec-extra-tables}

```{r}

bowling_batting_role_matchup_boundaries <- cleaned_data %>%
  group_by(bowling_style, batter_playing_role) %>%
  summarise(
    wickets_percentage = round(sum(wicket == TRUE)/n(), 3),  
    num_balls = n(),                             
  ) %>% arrange(desc(wickets_percentage), bowling_style, batter_playing_role)

bowling_batting_role_matchup_boundaries

```

@@

# Cleaned Data Dictionary {#sec-cleaned-data-dictionary}

@@

# Idealized Methodology

@@

## Translating Plays to Data

@@

## Making Data Easy to Use

@@

## Making Data Accessible

@@

# Data Sheet for The Raw Data {#sec-data-sheet}

## Questions

### Who put the data set together?

@@

### Who paid for the data set to be created?

@@

### How complete is the data set? {#sec-complete-data set}

@@

### Which variables are present, and, equally, not present, for particular observations?

@@

## Data Dictionary For Raw Data {#sec-data-dictionary}

@@

### Play-by-play IPL Data

@@

Season (String)  
Example: 2007/8 or 2024  
Represents the IPL tournament year of the match. For instance “2007/8” represents the 2007 to 2008 IPL season and 2024 represents the tournament that was played only in 2024\.

@@

### Player Metadata

@@

What to cite:
- cricketdata
- ESPNCricinfo
- Cricsheet
- All tidyverse packages used

>>>>

library(tidyverse)
library(arrow)
library(here)
library(knitr)
library(modelsummary)
library(marginaleffects)

@@

All cricket data is from the [@citeR] package [@cricketdata]. The data that was used is from the CRAN version of the package and not the [@github] developer tools version because the CRAN is more stable. [@cricketdata] takes data from [@espncricinfo] and [@cricsheet]. While not stated by [@cricketdata], based on the code the team behind the package are scrapping the HTML of ESPN Cricinfo and Cricsheet, As a result, anyone using the package must respect the rate limits set by these sites. Therefore, the data was first saved as CSV before performing any other cleaning and analysis.

More information about the raw data sets, measurements, [@cricsheet], [@espncricinfo], and [@cricketdata] is in @sec-data-sheet.

@@

```{r}
#| label: fig-prevoverwickets-wickets
#| fig-cap: Positive Association Between Wickets in the Last and Next Over
#| echo: false
#| warning: false
wickets_prev_over_wickets <- cleaned_data %>%
  group_by(over) %>%
  summarise(
    num_wickets = sum(wicket == TRUE),
    prev_over_wickets = mean(prev_over_wickets),
    num_balls = n(),                             
  ) %>% arrange(desc(num_wickets), desc(num_balls))

@@

## Model Set-up

All models are set up using the `glm` [@citeR] CRAN package. It uses the binomial family. All models trained on a training dataset, which contains 80% of the original data, which the other 20% being the testing set.

### Single Variable Generalized Linear Model

The single variable model uses a GLM model with only the over number as the input variable.

@@

Beta Explanation:

$\beta_0$: Represents the general intercept, which represents the log-odds of a wicket occurring when the over variables does not matter

$\beta_1$: Represents the coefficient for the current over number

@@

### Model Set-Up for Two Variable Variable Generalized Linear Model

The single variable model uses a GLM model with the over number and number of wickets before an out as the input variable.

@@

$\beta_1$: Represents the coefficient for the current over number

$\beta_2$: Represents the coefficient for the number of wickets in the previous over

@@

complex_glm_wicket_model <- readRDS(here("models/complex_glm_wicket_model.rds"))
#summary(complex_glm_wicket_model)
modelsummary(list("Wicket Prediction (Two Variable Input)" = complex_glm_wicket_model))
```
Based on @tbl-model-summary-two-var, modeling with only the over variable accounts 0.22 RSME variability. Also for an increase in number previous wickets in the previous over, the log odds of a wicket occurring increase by 1.852 with a p value that is almost zero (less than $2*10^-16$). The null deviance is 20849 on 51586 degrees of freedom while the residual deviance is 16584 on 51584 degrees of freedom. The logs odds increase for over is 0.005 with a p value of 0.168.

### Model Set-Up for Four Variable Generalized Linear Model

Adding the two extra variables, batting style and bowling style, to the model did not change the effectiveness in predictability at all. More rigorous analysis available in @sec-fourvar-model.

@@

Based on @fig-over-wickets, there was a strong positive linear relationship between the over and the number of wickets taken, suggesting that the over can be a predictor of a wicket occurring. From @fig-bowlingstyle-wickets certain match-up produced a higher chance of a wicket occurring, such as left hand fast on a left hand batter, which is why both variables were added to the four variable model in @sec-fourvar-model. Finally from @fig-prevoverwickets-wickets, the average number of wickets that occurred in the previous over had a strong positive linear relationship with the number of wickets that occurred in balls in the next over, implying that previous over wickets impact the wicket probability of a ball in the next over.

@@

@@

simple_glm_wicket_model_predictions |>
  mutate(wicket = factor(wicket)) |>
  ggplot(aes(x = over, y = estimate, color = wicket)) +
  stat_ecdf(geom = "point", alpha = 0.75) +
  labs(
    x = "The Over",
    y = "Estimated Probability that a wicket will occur",
    color = "Was actually a wicket"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```
Based on @fig-onevarmodel-results, there's a strong linear positive relationship between the over number and the single variable model wicket prediction probability. Also, in the 20th over, the model will almost always predict a wicket.

```{r}
#| label: tbl-model-one-var-results
#| tbl-cap: "One Variable Model Underfits the Test Data"
#| echo: false
#| warning: false
test_data_simple <- test_data
predictions <- predict(simple_glm_wicket_model, newdata = test_data_simple, type = "response")

@@

summary_results <- test_data_simple %>% group_by(wicket) %>% 
  summarise(
  correct = sum(correct_prediction),
  incorrect = sum(!correct_prediction)
) %>% mutate(wicket = if_else(wicket, "Yes", "No"))

summary_results %>% kable(col.names = c("Was Actually A Wicket?", "Correct", "Incorrect"))

```
From @tbl-model-one-var-results, the model has is 95.000% accurate, with a 100% accuracy for correctly predicting a wicket. However, the model has 0% change of correctly predicting when a wicket will occur. For this model, it is assumed that when the estimated probability of a wicket occurring is greater than or equal to 50%, the model is predicting a wicket.

@@

```{r}
#| label: fig-twovarmodel-results
#| fig-cap: Logarithmic Relationship Between Wicket Probability and the previous overs wickets
#| echo: false
#| warning: false
complex_glm_wicket_model_predictions <-
  predictions(complex_glm_wicket_model) |>
  as_tibble()

complex_glm_wicket_model_predictions |>
  mutate(wicket = factor(wicket)) |>
  ggplot(aes(x = prev_over_wickets, y = estimate, color = wicket)) +
  stat_ecdf(geom = "point", alpha = 0.75) +
  labs(
    x = "The Number of Wickets Taken in the Previous Over",
    y = "Estimated Probability that a wicket will occur",
    color = "Was actually a wicket"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```
Based on @fig-twovarmodel-results, the number of wickets in the previous over had a positive log-like relationship with the estimated probability of a wicket occurring the two variable model. Also, when the previous over had two or more wickets, the model predicts 90% or more chance of a wicket occurring. It is important to note that this model also factors in variation from the current over of the game when generating predictions probability estimates.

```{r}
#| label: tbl-model-two-var-results
#| tbl-cap: "Two Variable Model Struggles to Predict Wickets in Test Data"
#| echo: false
#| warning: false
test_data_complex <- test_data
predictions <- predict(complex_glm_wicket_model, newdata = test_data_complex, type = "response")

test_data_complex$predicted_wicket_prob <- predictions
test_data_complex <- test_data_complex %>% 
  mutate(predicted_wicket = predicted_wicket_prob >= 0.5) %>% 
  mutate(correct_prediction = predicted_wicket == wicket)

summary_results <- test_data_complex %>% group_by(wicket) %>% 
  summarise(
  correct = sum(correct_prediction),
  incorrect = sum(!correct_prediction)
) %>% mutate(wicket = if_else(wicket, "Yes", "No"))

summary_results %>% kable(col.names = c("Was Actually A Wicket?", "Correct", "Incorrect"))
```
Via @tbl-model-two-var-results, the two variable model has a 94.952% accuracy on the test data. Wicket occurrences are predicted with a 2.640% accuracy and a wicket not occurring is predicted with a 99.804% accuracy. For this model, it is assumed that when the estimated probability of a wicket occurring is greater than or equal to 50%, the model is predicting a wicket.

@@

The four variable model is discussed in the results section since most of the variables did not have a large impact on the model. More information on that model in @sec-fourvar-model.

@@

## Variables that Lead to Wickets

## Variables that Do Not Lead to Wickets

## Model Discussion

## Next Steps, Weaknesses, and Limitations

@@

## Four Variable Model {#sec-fourvar-model}

@@

$\beta_1$: Represents the coefficient for the current over number

$\beta_2$: Represents the coefficient for the number of wickets in the previous over

$\beta_3$: Represents the coefficient for the bowling style. Each bowling style has its own coefficient.

$\beta_4$: Represents the coefficient for the batting style (left handed or right handed). Each batting style has its own coefficient.

@@

```{r}
#| label: tbl-model-summary-four-var
#| tbl-cap: "Batting Style vs Bowling Style Do Not Heavily Impact Wicket Potential"
#| echo: false
#| warning: false
overly_complex_glm_wicket_model <- readRDS(here("models/overly_complex_glm_wicket_model.rds"))
#summary(overly_complex_glm_wicket_model)
modelsummary(overly_complex_glm_wicket_model)
```
Based on @tbl-model-summary-four-var, none of the batting styles or bowling style coefficient are statistically significant all with p values greater than 10%. This implies that most likely these variables not great predictors of wicket creation. The number of wickets in the previous over still has a statistically significant impact and has an increase of 0.031446 log odds per an increase of one wicket in the previous over. The residual deviance is 16574 on 51571 degrees of freedom compared to the null of 20849 on 51586 degrees of freedom.

## Variables Not Part of Model

@@

@@

@@

## Cleaned Data Dictionary {#sec-cleaned-data-dictionary}

@@

## Idealized Methodology

@@

### Translating Plays to Data

@@

### Making Data Easy to Use

@@

### Making Data Accessible

@@

## Data Sheet for The Raw Data {#sec-data-sheet}

### Questions

#### Who put the data set together?

@@

#### Who paid for the data set to be created?

@@

#### How complete is the data set?

@@

#### Which variables are present, and, equally, not present, for particular observations?

@@

### Data Dictionary For Raw Data {#sec-data-dictionary}

@@

#### Play-by-play IPL Data

@@

Season (String)  
Example: 2007/8 or 2024  
Represents the IPL tournament year of the match. For instance “2007/8” represents the 2007 to 2008 IPL season and 2024 represents the tournament that was played only in 2024.

@@

#### Player Metadata

@@
